---
title: "Alpine marmot Dispersal Model: Analysis"
author: "Pierre Dupont"
date: "12/09/2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alpine marmot Dispersal Model: Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates the marmot dispersal capture-recapture model, as in "Estimation of age and sex-specific natal dispersal patterns of Alpine marmots from capture-recapture data." (Dupont *et al*, *submitted*). Specifically, we implement the two versions of the CR model used in the paper using nimble.


### Load libraries and data
```{r packages, message=FALSE, warning=FALSE}
rm(list = ls())
library(nimble, quietly = T)
library(jagsUI)
load("marmotDispersal_Input.RData")
```



### Prepare JAGS data
Now, we format the data for JAGS and prepare initial values.
```{r data}
## 1. JAGS data
data <- list( y = y,
             f = f,
             nind = dim(y)[1],
             n.occ = dim(y)[2],
             dist = dist,
             site = site, 
             prob.mat = prob.mat,
             dmax = dmax,
             sex = sex,
             age = age,
             help = help,
             A = max(na.omit(age)))

## 2. Function to generate compatible initial values for z
INITS <- function(CH,f,PSI.STATE,PSI.OBS){						
  ##1.Model informations	
  imax <- nrow(CH)				##number of individuals
  tmax <- ncol(CH)				##number of occasions												
  n.states <- dim(PSI.OBS)[2]		##number of states												
  n.events <- dim(PSI.OBS)[3]		##number of events
  
  ##2.Read matrices function
  get.states <- function(x,i){ 						##read matrix backwards and returns row number
    res <- NULL
    for (j in 1:nrow(x)){ 
      if (x[j,i]!=0) 
        res <- cbind(res,j)
    }
    return(res)
  }
  
  get.next.state <- function (x,i){
    res <- NULL
    for (j in 1:ncol(x)){
      if (x[i,j]!=0)
        res <- cbind(res,j)
    }
    return(res)
  }
  
  ##3.Sample the possible states
  z <- matrix(NA,imax,tmax)						
  for (i in (1:imax)){
    #print(i)
    E <- S <- TOT <- F <- T <- list(rep(NA,1))
    E[[tmax]] <- S[[tmax]] <- TOT[[tmax]] <- get.states(PSI.OBS[i,,,tmax],CH[i,tmax])
    z[i,f[i]] <- F[[f[i]]] <- T[[f[i]]] <- get.states(PSI.OBS[i,,,f[i]],CH[i,f[i]])	
    
    for (t in ((tmax-1):(f[i]))){
      E[[t]] <- get.states(PSI.OBS[i,,,t],CH[i,t])
      
      X <- rep(NA,1)
      for (x in 1:length(TOT[[t+1]])){
        y <- get.states(PSI.STATE[i,,,t],TOT[[t+1]][x])
        X <- c(X,y)
      }
      
      S[[t]] <- unique(X)
      TOT[[t]] <- intersect(S[[t]],E[[t]])
    }
    
    for (t in ((f[i]+1):tmax)){
      F[[t]] <- get.next.state(PSI.STATE[i,,,t],z[i,t-1])
      T[[t]] <- intersect(F[[t]],TOT[[t]])
      fstate <- sample (length(T[[t]]),1)
      z[i,t] <- T[[t]][fstate]
    }
  }  
  for (i in 1:nrow(z)){
    fi <- z[i,f[i]]
    z[i,1:f[i]] <- NA
    z[i,f[i]] <- fi
  }
  return(z)
}

state.mat <- array(NA,c(dim(y)[1],4,4,dim(y)[2]))
obs.mat <- array(NA,c(dim(y)[1],4,4,dim(y)[2]))
disp <- matrix(NA,dim(y)[1],dim(y)[2])
for (i in 1:dim(y)[1]){
  for (t in f[i]:dim(y)[2]){ 
    disp[i,t] <- ifelse(sum(prob.mat[site[i],,t]) <= 0, 0, 1)
    state.mat[i,,,t] <- matrix(c(			#Transition matrix
      1, 1, disp[i,t], 1, 
      0, 1, 0, 1, 
      0, 0, 1, 1, 
      0, 0, 0, 1
    ), nrow = 4, byrow = TRUE)    
    obs.mat[i,,,t] <- matrix(c(			##Observation matrix	
      1, 0, 0, 1, 
      0, 1, 0, 1, 	
      0, 0, 1, 1, 
      0, 0, 0, 1                               
    ), nrow = 4, byrow = TRUE)
  }
}
jagsInits <- function(){list(z = INITS(y,f,state.mat,obs.mat))}

## 3. Parameters monitored
params_POIS <- c("PHIS",
                 "PHILB",
                 "DISP",
                 "HERIT",
                 "P",
                 "d.P",
                 "d.trois",
                 "d.quatre",
                 "tau")

```


### Define JAGS models
Now, we define the Natal & Apparent Dispersal models in JAGS language.
```{r JAGS models}
sink("natalDispModel_AlpineMarmots.jags")
cat("model {
  # Priors and constraints
  tauH ~ dgamma(0.01,0.01)I(0.00001,100)
  tau[1] ~ dgamma(0.01,0.01)I(0.00001,100)
  tau[2] ~ dgamma(0.01,0.01)I(0.00001,100)
  
  for (i in 1:nind){
    help[i] ~ dpois(tauH)
    dist[i] ~ dpois(tau[sex[i]])
    DD[i] <- step(dmax -(dist[i]+1))
    D[i] <- DD[i]+1
    DIST[i,1] <- dmax
    DIST[i,2] <- dist[i]+1
    for (t in f[i]:(n.occ-1)){
      phiS[i,t] <- mean.phiS[age[i,t],i]
      phiDB[i,t] <- mean.phiDB[age[i,t],sex[i]] 
      phiLB[i,t] <- mean.phiLB[age[i,t],sex[i]]  
      d[i,t] <- mean.d[age[i,t],sex[i]]
      h[i,t] <- mean.h[age[i,t],sex[i]]
      p[i,t] <- mean.p[age[i,t+1],sex[i],t]
      r[i,t] <- prob.mat[site[i],DIST[i,D[i]],t]
    } 
  }

  for (a in 1:A){
    PHIS[a,1] ~ dnorm(0,0.001)
    PHIDB[a,1] ~ dnorm(0,0.001)
    PHILB[a,1] ~ dnorm(0,0.001)
    DISP[a,1] ~ dnorm(0,0.001)
    HERIT[a,1] ~ dnorm(0,0.001)
    PHIS[a,2] ~ dnorm(0,0.001)
    PHIDB[a,2] ~ dnorm(0,0.001)
    PHILB[a,2] ~ dnorm(0,0.001)
    DISP[a,2] ~ dnorm(0,0.001)
    HERIT[a,2] ~ dnorm(0,0.001)
    logit(mean.phiDB[a,1]) <- PHIDB[a,1]
    logit(mean.phiDB[a,2]) <- PHIDB[a,2]
    logit(mean.phiLB[a,1]) <- PHILB[a,1]
    logit(mean.phiLB[a,2]) <- PHILB[a,2]
    logit(mean.d[a,1]) <- DISP[a,1]
    logit(mean.d[a,2]) <- DISP[a,2]
    logit(mean.h[a,1]) <- HERIT[a,1]
    logit(mean.h[a,2]) <- HERIT[a,2]
  }
  
    beta.H ~ dnorm(0,0.001)
  for(i in 1:nind){
    HELP[i] <- log(help[i]+1)
    logit(mean.phiS[1,i]) <- PHIS[1,sex[i]] + beta.H*HELP[i]
    for (a in 2:A){
      logit(mean.phiS[a,i]) <- PHIS[a,sex[i]]
    }
  }
  
  for (t in 1:(n.occ-1)){
    logit(mean.p[2,1,t]) <- P[t]
    logit(mean.p[2,2,t]) <- P[t]+d.P
    logit(mean.p[3,1,t]) <- P[t]+d.trois
    logit(mean.p[3,2,t]) <- P[t]+d.P+d.trois
    logit(mean.p[4,1,t]) <- P[t]+d.quatre
    logit(mean.p[4,2,t]) <- P[t]+d.P+d.quatre
    P[t] ~ dnorm(0,0.001)
  }
  d.trois ~ dnorm(0,0.001)
  d.quatre ~ dnorm(0,0.001)
  d.P ~ dnorm(0,0.001)
  
  # Define state-transition and observation matrices
  for (i in 1:nind){
    for (t in f[i]:(n.occ-1)){         
      # Define probabilities of state S(t+1) given S(t)   
      ps[1,i,t,1] <- phiS[i,t] * (1-d[i,t]) * (1-h[i,t]) 
      ps[1,i,t,2] <- phiS[i,t] * (1-d[i,t]) * h[i,t]
      ps[1,i,t,3] <- phiS[i,t] * d[i,t] * r[i,t]
      ps[1,i,t,4] <- 1-phiS[i,t] + phiS[i,t]*d[i,t]*(1-r[i,t])
      ps[2,i,t,1] <- 0
      ps[2,i,t,2] <- phiDB[i,t] 
      ps[2,i,t,3] <- 0
      ps[2,i,t,4] <- (1-phiDB[i,t]) 
      ps[3,i,t,1] <- 0
      ps[3,i,t,2] <- 0
      ps[3,i,t,3] <- phiLB[i,t]
      ps[3,i,t,4] <- (1-phiLB[i,t])
      ps[4,i,t,1] <- 0
      ps[4,i,t,2] <- 0
      ps[4,i,t,3] <- 0
      ps[4,i,t,4] <- 1
      
      # Define probabilities of O(t) given S(t)
      po[1,i,t,1] <- p[i,t]
      po[1,i,t,2] <- 0
      po[1,i,t,3] <- 0
      po[1,i,t,4] <- 1-p[i,t]
      po[2,i,t,1] <- 0
      po[2,i,t,2] <- p[i,t]
      po[2,i,t,3] <- 0
      po[2,i,t,4] <- 1-p[i,t]
      po[3,i,t,1] <- 0
      po[3,i,t,2] <- 0
      po[3,i,t,3] <- p[i,t]
      po[3,i,t,4] <- 1-p[i,t]
      po[4,i,t,1] <- 0
      po[4,i,t,2] <- 0
      po[4,i,t,3] <- 0
      po[4,i,t,4] <- 1
    } 
  } 
  
  # Likelihood 
  for (i in 1:nind){
    z[i,f[i]] ~ dcat(y[i,f[i]])
    for (t in (f[i]+1):n.occ){
      z[i,t] ~ dcat(ps[z[i,t-1], i, t-1, ])
      y[i,t] ~ dcat(po[z[i,t], i, t-1, ])
    } 
  } 
}
",fill = TRUE)
sink()
```


### JAGS model fitting
Then, we organize the simulated data in a format useable by 'JAGS' and fit the model using the jagsUI package.
```{r JAGS run}
jagsOutput <- jags( data,
                    jagsInits,
                    params_POIS,
                    "natalDispModel_AlpineMarmots.jags",
                    n.chains = 1,
                    n.thin = 1,
                    n.iter = 6000,
                    n.burnin = 1000)


jagsOutput$summary
plot(jagsOutput)
```

